---
title: "Multivariate Analysis of Scotch Whisky Components"
format: pdf
editor: visual
---

To Cover: - A high-level description of the study and its motivation. - A quantitative descriptive analysis of the data. - A graphical/qualitative analysis of the data. - Can you assume that the data follow a multivariate normal distribution? - no - Are there “surprising” observations? You must try to answer the question (with a proper discussion): Is the cluster analysis consistent with different distances and linkages?

## A Qualitative Analysis of Data

```{r}
#| label: load-packages
#| include: false
  
library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(gridExtra)
library(reshape2)
library(dplyr)
```

### Prepping data:

```{r}
whiskey.data <- read.csv("../whisky_trace_elements.csv", header=TRUE)

handle_lod <- function(column) {
  for (i in 1:length(column)) {
    val = column[i]
    if (grepl('<', val)) {
      new_val = as.double(sub("<","",val)) / 2.
      column[i] = new_val
    }
    column[i] = as.double(column[i])
  }
  return(column)
}

## handling lod's like the paper does
whiskey.data$P <- as.double(handle_lod(whiskey.data$P))
whiskey.data$Cl <- as.double(handle_lod(whiskey.data$Cl))
whiskey.data$Br <- as.double(handle_lod(whiskey.data$Br))
whiskey.data$Rb <- as.double(handle_lod(whiskey.data$Rb))

whiskey.data <- whiskey.data[, -1] # drop the sample no - technically implicitely exists cause indexing

whiskey.data$Descriptor <- as.factor(whiskey.data$Descriptor)
whiskey.data$Distillery <- as.factor(whiskey.data$Distillery)
whiskey.data <- as.data.frame(whiskey.data)

```

```{r}
whiskey.num <- whiskey.data[,-(1:2)] #dropping factors 
summary(whiskey.data[,-(1:2)])
```

Notes about data: - All values are above 0 - They aren't all in the same order of magnitude range, .e.g, Br is smaller than P.

### Plotting:

```{r}
# function for getting plots, so that I don't have to rewrite it :)
get.ggcorrplot <- function(data, title){
  plot <- ggcorrplot(data,
    method = "circle",
    hc.order = TRUE,
    type = "lower",
    title = title
    )
  plot
}
```

First looking at box plots of elements across Descriptor:

```{r}
whiskey.melt <- melt(whiskey.data[,-2], 
                     id.vars="Descriptor", 
                     variable.name = "Element")
max.across.all <- max(whiskey.melt$value)
uniq.desc <- unique(whiskey.data$Descriptor)

subsect.1.melt <- subset(whiskey.melt, Descriptor %in% c("Blend","Counterfeit","Grain"))
subsect.2.melt <- subset(whiskey.melt, Descriptor %in% c("Highland","Island","Lowland"))
subsect.3.melt <- subset(whiskey.melt, Descriptor == "Speyside")

get.boxplots <- function(data){
  ggplot(data=data, aes(x=Element, y=value)) +
    geom_boxplot(aes(col=Element), notch = F) +
    facet_grid(~ Descriptor) +
    ylab("Trace Amount") +
    ylim(0, max.across.all + max.across.all/20.) +
    theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
    legend.position = "none")
}

plots <- lapply(1:7, function(i) {
  get.boxplots(subset(whiskey.melt, Descriptor == uniq.desc[i]))
})
#row.1 <- get.boxplots(subsect.1.melt)
#row.2 <- get.boxplots(subsect.2.melt)
#row.3 <- get.boxplots(subsect.3.melt)
#test <- c(row.1, row.2, row.3)
#grid.arrange(grobs = plots, nrow = 3, heights=rep(4,3), widths=rep(2,3))
#grid.arrange(grobs = plots, layout_matrix = matrix(c(rep(1:3,3), rep(4:6, 3), rep(7:9, 3)), ncol = 3, byrow=T))
```

```{r}
get.boxplots(subsect.1.melt)
get.boxplots(subsect.2.melt)
get.boxplots(subsect.3.melt)
```

Looking at the boxplots there seem to be some differences. Counterfeit stans out a lot because of the hiigh amount of S it has. Island stands out for K's variance.

As the sample size is tiny, any attempt at adding notches causes it to go outside the range of the hinges (outside of the IQR). For an example:

```{r}
ggplot(data=subsect.3.melt, aes(x=Element, y=value)) +
  geom_boxplot(aes(col=Element), notch = T) +
  facet_grid(~ Descriptor) +
  ylab("Trace Amount") +
  theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
  legend.position = "none")
```

```{r}
get.ggcorrplot(cor(whiskey.data[,-(1:2)]), "Correlation Across all Classes")
```

```{r}
counterfit.corrplot <- get.ggcorrplot(cor(subset(whiskey.data, Descriptor=="Counterfeit")[,-(1:2)]), "Correlation for Counterfeit Whiskey")

non.counterfit.corrplot <- get.ggcorrplot(cor(subset(whiskey.data, Descriptor!="Counterfeit")[,-(1:2)]), "Correlation for Non-Counterfeit Whiskey")

grid.arrange(counterfit.corrplot, non.counterfit.corrplot, ncol=2)
```

There's a clear indicator for some difference with a lot more negative correlations for the counterfeit whiskeys compared to the others.

**Want to check next for:** - Look at how it differs if filtering for specific descriptors. - Outliers? (or odd looking measurements) - proper understanding of how to deal with those '\<xyz' measurements - have ggcorrplot maintain label ordering for easier comparing...

### Mahalanobis:

```{r}
cov(whiskey.num)
```

```{r}
p <- 11
dM <- mahalanobis(whiskey.data[-(1:2)], center=colMeans(whiskey.data[-(1:2)]), cov=cov(whiskey.data[-(1:2)]))

upper.quantiles <- qchisq(c(.9, .95, .99), df=p)
density.at.quantiles <- dchisq(x=upper.quantiles, df=p)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)

ggplot(data.frame(dM), aes(x=dM)) +
	geom_histogram(aes(y=after_stat(density)), bins=8,
				   fill="white", col="black") +
	geom_rug() +
	stat_function(fun=dchisq, args = list(df=p),
				  col="red", size=2, alpha=.7, xlim=c(-5,40)) +
	geom_segment(data=cut.points,
				 aes(x=upper.quantiles, xend=upper.quantiles,
					 y=rep(0,3), yend=density.at.quantiles),
				 col="blue", size=2) +
	xlab("Mahalanobis distances and cut points") +
	ylab("Histogram and density")
```

### PCA

```{r}
whiskey.log <- whiskey.data |> mutate(across(where(is.numeric), log))
whiskey.stand <- whiskey.log |> mutate(across(where(is.numeric), ~(scale(.) %>% as.vector)))

#ggplot(whiskey.stand, aes(x=Zn)) +
#  geom_histogram(bins=30, fill="white", col="black")

#summary(whiskey.stand)



(Sigma <- cov(whiskey.stand[3:13]))
eigen(Sigma)
pca <- as.matrix(whiskey.stand[3:13]) %*% eigen(Sigma)$vectors
```

```{r}
pca <- as.data.frame(pca)
pca$Descriptor <- whiskey.data$Descriptor
summary(pca)

ggplot(pca, aes(x=V1,y=V2, color=factor(Descriptor)))+
  geom_point(size=1)
```

```{r}
#whiskey.stand$Descriptor <- whiskey.data$Descriptor
#whiskey.stand.melt <- melt(whiskey.stand, 
#                           id.vars="Descriptor", 
#                           variable.name = "Element")
#whiskey.stand.melt

#ggplot(whiskey.stand.melt, aes(x=value))+
#  geom_histogram(bins=nclass.FD(whiskey.stand.melt$value))+
#  geom_rug()
```

Looks closer to a poison or chi squared?

## Normal Assumptions

```{r}
plot_norm_est <- function(data, parameter, t) {
  ggplot(data) +
    geom_histogram(mapping = aes(x = parameter, y=..density..), fill="steelblue", colour="black", bins = nclass.FD(parameter)) +
    stat_function(fun=dnorm, args= list(mean=0, sd=1), xlim=c(-4, 4)) + 
    ggtitle(t)
}

plot_norm_est(whiskey.stand, whiskey.stand$P, "P")
plot_norm_est(whiskey.stand, whiskey.stand$S, "S")
plot_norm_est(whiskey.stand, whiskey.stand$Cl, "Cl")
plot_norm_est(whiskey.stand, whiskey.stand$K, "K")
plot_norm_est(whiskey.stand, whiskey.stand$Ca, "Ca")
plot_norm_est(whiskey.stand, whiskey.stand$Mn, "Mn")
plot_norm_est(whiskey.stand, whiskey.stand$Fe, "Fe")
plot_norm_est(whiskey.stand, whiskey.stand$Cu, "Cu")
plot_norm_est(whiskey.stand, whiskey.stand$Zn, "Zn")
plot_norm_est(whiskey.stand, whiskey.stand$Br, "Br")
plot_norm_est(whiskey.stand, whiskey.stand$Rb, "Rb")

```

Hardly any of these look normal

```{r}
plot.qq <- function(data, parameter, title) {
  ggplot(data, aes(sample=parameter)) +
    stat_qq(mapping = aes(colour = Descriptor)) + stat_qq_line()+
    ggtitle(paste("Normal QQ Plot for", title, sep=" "))
}
plot.qq(whiskey.stand, whiskey.stand$P, "P")
plot.qq(whiskey.stand, whiskey.stand$S, "S")
plot.qq(whiskey.stand, whiskey.stand$Cl, "Cl")
plot.qq(whiskey.stand, whiskey.stand$K, "K")
plot.qq(whiskey.stand, whiskey.stand$Ca, "Ca")
plot.qq(whiskey.stand, whiskey.stand$Mn, "Mn")
plot.qq(whiskey.stand, whiskey.stand$Fe, "Fe")
plot.qq(whiskey.stand, whiskey.stand$Cu, "Cu")
plot.qq(whiskey.stand, whiskey.stand$Zn, "Zn")
plot.qq(whiskey.stand, whiskey.stand$Br, "Br")
plot.qq(whiskey.stand, whiskey.stand$Rb, "Rb")

```

Again, all of these look skewed to the right. Fe looks the closest

Interestingly you can make out the counterfeits seem to be the strange observations in S.

## New Conclusion About normality

After actually reading the paper they mention log transforming the data. Doing that makes the distributions look normal T-T lol PCA also looks like what they got

## Classifiying by distance

```{r}
library(dplyr)
p <- 11
whiskey.log.counter <- whiskey.log |> filter(Descriptor == "Counterfeit")
whiskey.log.no.counter <- whiskey.log |> filter(Descriptor != "Counterfeit")
qr(whiskey.log.counter[-(1:2)])$rank
#dM.counter <- mahalanobis(whiskey.log.counter[-(1:2)], center=colMeans(whiskey.log.counter[-(1:2)]), cov=cov(whiskey.log.counter[-(1:2)]))
#dM.non.counter <- mahalanobis(whiskey.log.no.counter[-(1:2)], center=colMeans(whiskey.log.no.counter[-(1:2)]), cov=cov(whiskey.log.no.counter[-(1:2)]))
```

Can't call the above code because counter has less observation than parameters, so trying with first 2 pca components

```{r}
pca.counter <- pca |> filter(Descriptor == "Counterfeit")
pca.no.counter <- pca |> filter(Descriptor != "Counterfeit")

X1 <- pca.counter[(1:2)]
X2 <- pca.no.counter[(1:2)]
p <- 2

dM.counter <- mahalanobis(X1, center=colMeans(X1), cov=cov(X1))
dM.non.counter <- mahalanobis(X1, center=colMeans(X2), cov=cov(X2))
```

Comparing counterfeit observations to models

```{r}
pchisq(dM.counter, df=p, lower.tail=FALSE)
pchisq(dM.non.counter, df=p, lower.tail=FALSE)
```

All counterfeit observations are closer to a a multivariate normal with just counterfeits compared to multivariate normal without counterfeit obsevations.

```{r}
dM.counter <- mahalanobis(X2, center=colMeans(X1), cov=cov(X1))
dM.non.counter <- mahalanobis(X2, center=colMeans(X2), cov=cov(X2))
counter <- pchisq(dM.counter, df=p, lower.tail=FALSE)
non.counter <- pchisq(dM.non.counter, df=p, lower.tail=FALSE)
data.frame(counter, non.counter, non.counter > counter)
```

All non-counterfeit ones get correctly classified by distance

TODO NEXT: - drop the value i'm test....
