---
title: "Normal/Mahalanobis and Outliers"
format: html
---

```{r}
#| label: Load Libraries
#| include: false

library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(gridExtra)
library(reshape2)
library(dplyr)
library(pander)
library(ggbiplot)
library(caret)
```

```{r}
#| label: Set Theme
#| include: false

set_theme(
  theme_bw(
    base_size = 12,
    base_family = "sans"
  ) +
    theme(
      panel.grid.major = element_line(colour = "grey85"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)

descriptor_colours <- c(
  "Blend"       = "#E41A1C",
  "Counterfeit" = "#377EB8",
  "Grain"       = "#4DAF4A",
  "Highland"    = "#984EA3",
  "Island"      = "#FF7F00",
  "Lowland"     = "#A65628",
  "Speyside"    = "#F781BF"
)

descriptor_shapes <- c(
  "Blend"       = 16,
  "Counterfeit" = 17,
  "Grain"       = 15,
  "Highland"    = 18,
  "Island"      = 3,
  "Lowland"     = 7,
  "Speyside"    = 8
)

## Want some kinda colours for bar/histograms

histo.col.1 <- "#6ea81c" 
histo.col.2 <- "#c444b3"
histo.outline <- "#4f4f4f"
```

```{r}
#| label: Prepping Data
#| include: false

whisky.data <- read.csv("../whisky_trace_elements.csv", header=TRUE)

elements <- c("P", "S", "Cl", "K", "Ca", "Mn",
              "Fe", "Cu", "Zn", "Br", "Rb")

handle_lod <- function(column) {
  for (i in 1:length(column)) {
    val = column[i]
    if (grepl('<', val)) {
      new_val = as.double(sub("<","",val)) / 2.
      column[i] = new_val
    }
    column[i] = as.double(column[i])
  }
  return(column)
}

## handling lod's like the paper does
whisky.data$P <- as.double(handle_lod(whisky.data$P))
whisky.data$Cl <- as.double(handle_lod(whisky.data$Cl))
whisky.data$Br <- as.double(handle_lod(whisky.data$Br))
whisky.data$Rb <- as.double(handle_lod(whisky.data$Rb))

whisky.data$Descriptor <- as.factor(whisky.data$Descriptor)
whisky.data$Distillery <- as.factor(whisky.data$Distillery)
whisky.data <- as.data.frame(whisky.data)
```

## 7. Multivariate Normality 
Can we assume a multivariate normal distribution?

### 7.1. Marginal normality

#### Histograms

```{r}
#| label: Fig_Histogram_By_Element


## From Mio's Report
whisky.long <- whisky.data |>
  pivot_longer(cols = all_of(elements),
               names_to = "Element", values_to = "Concentration")

ggplot(whisky.long, aes(x = Concentration)) +
  geom_histogram(bins = 10, colour = histo.outline, fill = histo.col.1) +
  facet_wrap(~ Element, scales = "free", ncol = 4) +
  labs(x = "Concentration (mg/L)", y = "Frequency")
```

#### QQ Plots 

```{r}
#| label: Fig_QQ_By_Element
#| fig-width: 8
#| fig-height: 6

ggplot(whisky.long, aes(sample=Concentration)) +
  stat_qq(mapping = aes(colour = Descriptor), size= 1) + 
  stat_qq_line() + 
  facet_wrap(~ Element, scales = "free", ncol = 4) +
  labs(x = "Theoretical Quantile", y = "Sample Quantile") +
  scale_color_manual(values = descriptor_colours)
```

#### Raw variables

#### Log Transformed Variables

```{r}
whisky.log <- whisky.data |>
  mutate(across(where(is.numeric), log))

whisky.long.log <- whisky.log|>
  pivot_longer(cols = all_of(elements),
               names_to = "Element", values_to = "Concentration")
```

```{r}
#| label: Fig_Histograms_By_Element_Log

ggplot(whisky.long.log, aes(x = Concentration)) +
  geom_histogram(bins = nclass.FD(whisky.long.log$Concentration), colour = histo.outline, fill = histo.col.2) +
  facet_wrap(~ Element, scales = "free", ncol = 4) +
  labs(x = "Concentration (mg/L)", y = "Frequency")
```
```{r}
#| label: Fig_QQ_By_Element_Log
#| fig-width: 8
#| fig-height: 6

ggplot(whisky.long.log, aes(sample=Concentration)) +
  stat_qq(mapping = aes(colour = Descriptor), size= 1) + 
  stat_qq_line() + 
  facet_wrap(~ Element, scales = "free", ncol = 4) +
  labs(x = "Theoretical Quantile", y = "Sample Quantile") +
  scale_color_manual(values = descriptor_colours)
```
Does log transformation improve normality?
Are there remaining skewed variables?
Are the tails problematic?

### 7.2. Multivariate normality 


```{r}
p <- 11

dM <- mahalanobis(whisky.log[-(1:3)], center=colMeans(whisky.log[-(1:3)]), cov=cov(whisky.log[-(1:3)]))

dM <- as.data.frame(dM)
names(dM) <- c("distance")
dM$Descriptor <- whisky.data$Descriptor
dM$Sample_no <- whisky.data$Sample_no

## From https://stackoverflow.com/questions/64950718/in-r-ggplot-how-to-add-labels-to-points-in-a-stat-qq
myplot <- ggplot(dM, aes(sample=distance)) +
  stat_qq(mapping = aes(colour = Descriptor), size= 1, 
          distribution = stats::qchisq, dparams=list(df=p)) + 
  stat_qq_line(distribution = stats::qchisq, dparams = list(df=p)) +
  ylim(c(3,27)) # so that the sample number appears correctly

x.pnts <- ggplot_build(myplot)$data[[1]]$x
y.pnts <- ggplot_build(myplot)$data[[1]]$y
offset <- (max(y.pnts) - min(y.pnts)) / 20

myplot +
    geom_text(label = dM$Sample_no,
              x = x.pnts,
              y = y.pnts + offset, 
              alpha = .6)
```

### 7.3. Conclusion 

Multivariate normality is approximately reasonable but imperfect after log-transforming the original data.

why?

## 8. Surprising Observations 

Are there surprising observations?

look at other methods for outliers?
- Z-Score Method
- IQR
- MAD - mentioned in lectures, so might be worthwhile?

### 8.1. Mahalanobis distance 

```{r}

upper.quantiles <- qchisq(c(.9, .95, .99), df=p)
density.at.quantiles <- dchisq(x=upper.quantiles, df=p)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)

ggplot(dM, aes(x=distance)) +
	geom_histogram(aes(y=after_stat(density)), bins=10, col=histo.outline, fill="lightgray") +
	geom_rug() +
	stat_function(fun=dchisq, args = list(df=p),
				  col="blue", linewidth=2, alpha=.9, xlim=c(-5,40)) +
	geom_segment(data=cut.points,
				 aes(x=upper.quantiles, xend=upper.quantiles,
					 y=rep(0,3), yend=density.at.quantiles),
				 col="green", linewidth=2) +
	xlab("Mahalanobis distances and cut points") +
	ylab("Histogram and density")
```
There seems to be observation above the $99\%$ cut-off point.

```{r}
pander(dM |> slice_max(distance, n=5))
```

```{r}
pander(whisky.data |> filter(Sample_no %in% (dM |> slice_max(distance, n=5))$Sample_no))
```

### 8.2 Investigate unusual samples 

sample 12:
- P; close to average
- S; very high (not the highest tho)
- Cl; definitly to the right tail, but not that out-there
- K; close to average
- Ca; relatively close to average
- Mn; Not too sure...
- Fe; Not too sure...
- Cu; close-ish to average
- Zn; again, not *too* out of the ordinary
- Br; *veryy* high
- Rb; super low


How does it compare to other counterfeit variables


### 8.3. Important distinction 

