---
title: "Multivariate Analysis of Scotch Whisky Components"
format: html
editor: visual
---

To Cover:
• A high-level description of the study and its motivation.
• A quantitative descriptive analysis of the data.
• A graphical/qualitative analysis of the data.
• Can you assume that the data follow a multivariate normal distribution?
• Are there “surprising” observations?
You must try to answer the question (with a proper discussion):
• Is the cluster analysis consistent with different distances and linkages?

## A Qualitative Analysis of Data

```{r}
#| label: load-packages
#| include: false
  
library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(gridExtra)
```

### Prepping data:

```{r}
whiskey.data <- read.csv("../whisky_trace_elements.csv", header=TRUE)

## handling it pretty poorly rn - need to research how it's properly done
whiskey.data$P <- as.double(gsub("<","",whiskey.data$P))
whiskey.data$Cl <- as.double(gsub("<","",whiskey.data$Cl))
whiskey.data$Br <- as.double(gsub("<","",whiskey.data$Br))
whiskey.data$Rb <- as.double(gsub("<","",whiskey.data$Rb))

whiskey.data <- whiskey.data[, -1] # drop the sample no - technically implicitely exists cause indexing

whiskey.data$Descriptor <- as.factor(whiskey.data$Descriptor)
whiskey.data$Distillery <- as.factor(whiskey.data$Distillery)
whiskey.data <- as.data.frame(whiskey.data)
```


```{r}
summary(whiskey.data[,-(1:2)])
```
Notes about data:
- All values are above 0
- They aren't all in the same order of magnitude range, .e.g, Br is smaller than P.

### Plotting:

```{r}
# function for getting plots, so that I don't have to rewrite it :)
get.ggcorrplot <- function(data, title){
  plot <- ggcorrplot(data,
    method = "circle",
    hc.order = TRUE,
    type = "lower",
    title = title
    )
  plot
}
```

```{r}
get.ggcorrplot(cor(whiskey.data[,-(1:2)]), "Correlation Across all Classes")
```

```{r}
counterfit.corrplot <- get.ggcorrplot(cor(subset(whiskey.data, Descriptor=="Counterfeit")[,-(1:2)]), "Correlation for Counterfeit Whiskey")

non.counterfit.corrplot <- get.ggcorrplot(cor(subset(whiskey.data, Descriptor!="Counterfeit")[,-(1:2)]), "Correlation for Non-Counterfeit Whiskey")

grid.arrange(counterfit.corrplot, non.counterfit.corrplot, ncol=2)
```
There's a clear indicator for some difference with a lot more negative correlations for the counterfeit whiskeys compared to the others.

**Want to check next for:**
- Look at how it differs if  filtering for specific descriptors.
- Outliers? (or odd looking measurements)
- proper understanding of how to deal with those '<xyz' measurements
- have ggcorrplot maintain label ordering for easier comparing...


