---
title: "Multivariate Analysis of Scotch Whisky Components"
format: html
editor: visual
---

To Cover:
• A high-level description of the study and its motivation.
• A quantitative descriptive analysis of the data.
• A graphical/qualitative analysis of the data.
• Can you assume that the data follow a multivariate normal distribution?
• Are there “surprising” observations?
You must try to answer the question (with a proper discussion):
• Is the cluster analysis consistent with different distances and linkages?

## A Qualitative Analysis of Data

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(ggplot2)
library(ggcorrplot)
```

Prepping data:

```{r}
whiskey.data <- read.csv("../whisky_trace_elements.csv", header=TRUE)

## handling it pretty poorly rn - need to research how it's properly done
whiskey.data$P <- as.double(gsub("<","",whiskey.data$P))
whiskey.data$Cl <- as.double(gsub("<","",whiskey.data$Cl))
whiskey.data$Br <- as.double(gsub("<","",whiskey.data$Br))
whiskey.data$Rb <- as.double(gsub("<","",whiskey.data$Rb))

whiskey.data$Descriptor <- as.factor(whiskey.data$Descriptor)
whiskey.data$Distillery <- as.factor(whiskey.data$Distillery)
whiskey.data <- as.data.frame(whiskey.data)
head(whiskey.data)
```
Plotting:
```{r}
ggcorrplot(cor(whiskey.data[,-(1:3)]),
  method = "circle",
  hc.order = TRUE,
  type = "lower")
```

**Want to check next for:**
- Look at how it differs if  filtering for specific descriptors.
- Outliers? (or odd looking measurements)
- proper understanding of how to deal with those '<xyz' measurements


