---
title: "Multivariate Analysis of Scotch Whisky Components"
format: html
editor: visual
---

To Cover:
• A high-level description of the study and its motivation.
• A quantitative descriptive analysis of the data.
• A graphical/qualitative analysis of the data.
• Can you assume that the data follow a multivariate normal distribution?
• Are there “surprising” observations?
You must try to answer the question (with a proper discussion):
• Is the cluster analysis consistent with different distances and linkages?

## A Qualitative Analysis of Data

```{r}
#| label: load-packages
#| include: false
  
library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(gridExtra)
library(reshape2)
```

### Prepping data:

```{r}
whiskey.data <- read.csv("../whisky_trace_elements.csv", header=TRUE)

## handling it pretty poorly rn - need to research how it's properly done
whiskey.data$P <- as.double(gsub("<","",whiskey.data$P))
whiskey.data$Cl <- as.double(gsub("<","",whiskey.data$Cl))
whiskey.data$Br <- as.double(gsub("<","",whiskey.data$Br))
whiskey.data$Rb <- as.double(gsub("<","",whiskey.data$Rb))

whiskey.data <- whiskey.data[, -1] # drop the sample no - technically implicitely exists cause indexing

whiskey.data$Descriptor <- as.factor(whiskey.data$Descriptor)
whiskey.data$Distillery <- as.factor(whiskey.data$Distillery)
whiskey.data <- as.data.frame(whiskey.data)
```


```{r}
whiskey.num <- whiskey.data[,-(1:2)] #dropping factors 
summary(whiskey.data[,-(1:2)])
```
Notes about data:
- All values are above 0
- They aren't all in the same order of magnitude range, .e.g, Br is smaller than P.

### Plotting:

```{r}
# function for getting plots, so that I don't have to rewrite it :)
get.ggcorrplot <- function(data, title){
  plot <- ggcorrplot(data,
    method = "circle",
    hc.order = TRUE,
    type = "lower",
    title = title
    )
  plot
}
```

First looking at box plots of elements across Descriptor:
```{r}
whiskey.melt <- melt(whiskey.data[,-2], 
                     id.vars="Descriptor", 
                     variable.name = "Element")
max.across.all <- max(whiskey.melt$value)
uniq.desc <- unique(whiskey.data$Descriptor)

subsect.1.melt <- subset(whiskey.melt, Descriptor %in% c("Blend","Counterfeit","Grain"))
subsect.2.melt <- subset(whiskey.melt, Descriptor %in% c("Highland","Island","Lowland"))
subsect.3.melt <- subset(whiskey.melt, Descriptor == "Speyside")

get.boxplots <- function(data){
  ggplot(data=data, aes(x=Element, y=value)) +
    geom_boxplot(aes(col=Element), notch = F) +
    facet_grid(~ Descriptor) +
    ylab("Trace Amount") +
    ylim(0, max.across.all + max.across.all/20.) +
    theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
    legend.position = "none")
}

plots <- lapply(1:7, function(i) {
  get.boxplots(subset(whiskey.melt, Descriptor == uniq.desc[i]))
})
#row.1 <- get.boxplots(subsect.1.melt)
#row.2 <- get.boxplots(subsect.2.melt)
#row.3 <- get.boxplots(subsect.3.melt)
#test <- c(row.1, row.2, row.3)
#grid.arrange(grobs = plots, nrow = 3, heights=rep(4,3), widths=rep(2,3))
#grid.arrange(grobs = plots, layout_matrix = matrix(c(rep(1:3,3), rep(4:6, 3), rep(7:9, 3)), ncol = 3, byrow=T))
```
```{r}
get.boxplots(subsect.1.melt)
get.boxplots(subsect.2.melt)
get.boxplots(subsect.3.melt)
```
Looking at the boxplots there seem to be some differences. Counterfeit stans out a lot because of the hiigh amount of S it has. Island stands out for K's variance.

As the sample size is tiny, any attempt at adding notches causes it to go outside the range of the hinges (outside of the IQR).
For an example:
```{r}
ggplot(data=subsect.3.melt, aes(x=Element, y=value)) +
  geom_boxplot(aes(col=Element), notch = T) +
  facet_grid(~ Descriptor) +
  ylab("Trace Amount") +
  theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
  legend.position = "none")
```

```{r}
get.ggcorrplot(cor(whiskey.data[,-(1:2)]), "Correlation Across all Classes")
```

```{r}
counterfit.corrplot <- get.ggcorrplot(cor(subset(whiskey.data, Descriptor=="Counterfeit")[,-(1:2)]), "Correlation for Counterfeit Whiskey")

non.counterfit.corrplot <- get.ggcorrplot(cor(subset(whiskey.data, Descriptor!="Counterfeit")[,-(1:2)]), "Correlation for Non-Counterfeit Whiskey")

grid.arrange(counterfit.corrplot, non.counterfit.corrplot, ncol=2)
```

There's a clear indicator for some difference with a lot more negative correlations for the counterfeit whiskeys compared to the others.

**Want to check next for:**
- Look at how it differs if  filtering for specific descriptors.
- Outliers? (or odd looking measurements)
- proper understanding of how to deal with those '<xyz' measurements
- have ggcorrplot maintain label ordering for easier comparing...

### Mahalanobis:

```{r}
cov(whiskey.num)
```

```{r}
p <- 11
dM <- mahalanobis(whiskey.num, center=colMeans(whiskey.num), cov=cov(whiskey.num))

upper.quantiles <- qchisq(c(.9, .95, .99), df=p)
density.at.quantiles <- dchisq(x=upper.quantiles, df=p)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)

ggplot(data.frame(dM), aes(x=dM)) +
	geom_histogram(aes(y=after_stat(density)), bins=8,
				   fill="white", col="black") +
	geom_rug() +
	stat_function(fun=dchisq, args = list(df=p),
				  col="red", size=2, alpha=.7, xlim=c(-5,40)) +
	geom_segment(data=cut.points,
				 aes(x=upper.quantiles, xend=upper.quantiles,
					 y=rep(0,3), yend=density.at.quantiles),
				 col="blue", size=2) +
	xlab("Mahalanobis distances and cut points") +
	ylab("Histogram and density")
```

### PCA