---
title: "Multivariate Analysis of Scotch Whisky Components"
format: html
editor: source
---

```{r}
#| label: load-packages
#| include: false
## Setup  
library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(gridExtra)
library(reshape2)
library(dplyr)
library(pander)
library(ggbiplot)
library(caret)
```

To Cover: - A high-level description of the study and its motivation. - A quantitative descriptive analysis of the data. - A graphical/qualitative analysis of the data. - Can you assume that the data follow a multivariate normal distribution? - no - Are there “surprising” observations? You must try to answer the question (with a proper discussion): Is the cluster analysis consistent with different distances and linkages?


## Abstract

## Introduction

## Data

The data contains contains 32 observations across:
- Counterfeit
- Blend
- Grain
- Highland
- Lowland
- Speyside

With trace elements for, (P, S, Cl, K, Ca, Mn, Fe, Cu, Zn, Br and Rb), measured  in mg/L.

12 out of the 32 observations have at least one of the trace elements detected below LOD.
All the counterfeit whiskey observations also have at least one variables below LOD

Amount of observations below LOD by individual element:
```{r}
#| echo: false

whiskey.data <- read.csv("../whisky_trace_elements.csv", header=TRUE)
whiskey.data <- as.data.frame(whiskey.data)

pander((whiskey.data |>
  summarise(
    across(everything(), ~ sum(startsWith(as.character(.), "<"), na.rm = TRUE))
  ))[-(1:3)])
```

### Transformations on Data
#### Handling the LOD values

If we were to drop observations with trace elements that were measured below LOD, we'd lose more than a third of the dataset, including all counterfeit observations, because of this they will instead be substitued.

The two common ways of substituting LOD values are, $LOD/2$ and $LOD/\sqrt{2}$. Similar to the paper that this data is from, this report will be subtituing all LOD values with $LOD/2$.

```{r}
#| echo: false
whiskey.data <- read.csv("../whisky_trace_elements.csv", header=TRUE)

handle_lod <- function(column) {
  for (i in 1:length(column)) {
    val = column[i]
    if (grepl('<', val)) {
      new_val = as.double(sub("<","",val)) / 2.
      column[i] = new_val
    }
    column[i] = as.double(column[i])
  }
  return(column)
}

## handling lod's like the paper does
whiskey.data$P <- as.double(handle_lod(whiskey.data$P))
whiskey.data$Cl <- as.double(handle_lod(whiskey.data$Cl))
whiskey.data$Br <- as.double(handle_lod(whiskey.data$Br))
whiskey.data$Rb <- as.double(handle_lod(whiskey.data$Rb))

whiskey.data <- whiskey.data[, -1] 

whiskey.data$Descriptor <- as.factor(whiskey.data$Descriptor)
whiskey.data$Distillery <- as.factor(whiskey.data$Distillery)
whiskey.data <- as.data.frame(whiskey.data)

whiskey.num <- whiskey.data[,-(1:3)] #dropping factors + sample number 
summary(whiskey.data[,-(1:3)])
```

From the summary we can see all values are above 0, but they aren't all in the same magnitured of range. E.g., Br is a lot smaller than P.

#### Log Transform on the data

Lastly, looking at (standardized) histogram of all the variables, we can see it looks rather skew,
```{r}
whiskey.stand <- whiskey.data |> 
  mutate(across(where(is.numeric), ~(scale(.) %>% as.vector)))
whiskey.stand$Descriptor <- whiskey.data$Descriptor
whiskey.stand.melt <- melt(whiskey.stand[-2], 
                           id.vars="Descriptor", 
                           variable.name = "Element")

ggplot(whiskey.stand.melt, aes(x=value))+
  geom_histogram(bins=nclass.FD(whiskey.stand.melt$value))+
  geom_rug() + 
  ggtitle("standardized histogram of trace elements")
```
After applying a log transform, the data looks a lot closer to a normal distribution.
```{r}
whiskey.stand <- whiskey.data |> 
  mutate(across(where(is.numeric), log)) |>
  mutate(across(where(is.numeric), ~(scale(.) %>% as.vector))) 
whiskey.stand.melt <- melt(whiskey.stand[-2], 
                           id.vars="Descriptor", 
                           variable.name = "Element")

ggplot(whiskey.stand.melt, aes(x=value))+
  geom_histogram(aes(y = ..density..), bins=nclass.FD(whiskey.stand.melt$value))+
  stat_function(fun = dnorm, args = list(mean=0, sd=1), color="red") +
  geom_rug() +
  ggtitle("log transformed standardized histogram of trace elements")
```

## A Qualitative Analysis of Data


### Box Plots

First looking at box plots of elements across Descriptor:

```{r}
whiskey.melt <- melt(whiskey.data[,-2], 
                     id.vars="Descriptor", 
                     variable.name = "Element")
uniq.desc <- unique(whiskey.data$Descriptor)

get.boxplots <- function(data, y_min, y_max){
  ggplot(data=data, aes(x=Element, y=value)) +
    geom_boxplot(aes(col=Element), notch = F) +
    facet_grid(~ Descriptor) +
    ylab("Trace Amount") +
    ylim(y_min, y_max) +
    theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
    legend.position = "none")
}

plots <- lapply(1:7, function(i) {
  get.boxplots(subset(whiskey.melt, Descriptor == uniq.desc[i]), 0, max(whiskey.melt$value))
})

grid.arrange(grobs = plots, nrow = 3, main="log transformed Trace Element based on Descriptor")
```

Because K's magnitude is so great compared to the other trace elements, standerdizing the data helps better compare across all trace elements.

```{r}
whiskey.stand.melt <- melt(whiskey.stand[,-2], 
                     id.vars="Descriptor", 
                     variable.name = "Element")

plots <- lapply(1:7, function(i) {
  get.boxplots(subset(whiskey.stand.melt, Descriptor == uniq.desc[i]), -3.5, 3.5)
})

grid.arrange(grobs = plots, nrow = 3, main="log transformed and stranderdized Trace Element based on Descriptor")
```

Looking at the boxplots there seem to be some differences. Counterfeit stands out a lot because of the high amount of S it compared to the others. Lowland has a suprisingly little amount of variance across all of it's elements.

#### Note on Notches for box plots
As the sample size is tiny, any attempt at adding notches causes it to go outside the range of the hinges (outside of the IQR). For an example:

```{r}
ggplot(data=subsect.3.melt, aes(x=Element, y=value)) +
  geom_boxplot(aes(col=Element), notch = T) +
  facet_grid(~ Descriptor) +
  ylab("Trace Amount") +
  theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
  legend.position = "none")
```

### Correlograms

```{r}
#| include: false
# function for getting plots, so that I don't have to rewrite it :)
get.ggcorrplot <- function(data, title){
  plot <- ggcorrplot(data,
    method = "circle",
    hc.order = FALSE,
    type = "lower",
    title = title,
    lab = TRUE,
    lab_size =2.5
    )
  plot
}
```

```{r}
get.ggcorrplot(cor(whiskey.data[,-(1:2)]), "Correlation Across all Classes")
```
**K**, **P**, and **RB** all have strong postiive correlations with one another, 

```{r}
counterfit.corrplot <- get.ggcorrplot(cor(subset(whiskey.data, Descriptor=="Counterfeit")[,-(1:2)]), "Correlation for Counterfeit Whiskey")

non.counterfit.corrplot <- get.ggcorrplot(cor(subset(whiskey.data, Descriptor!="Counterfeit")[,-(1:2)]), "Correlation for Non-Counterfeit Whiskey")

grid.arrange(counterfit.corrplot, non.counterfit.corrplot, ncol=2)
```

There's a clear indicator for some difference with a lot more negative correlations for the counterfeit whiskeys compared to the others. Particularly along **Fe**.
Although, the strongest correlations are still those positive ones.

When separating out based on Counterfeit vs Non-Counterfeit, Non-Counterfeit still matains those strong correlations, with the greatest still being that between, **K**, **P**, and **RB**. 
This might be a common trace element picked up during whiskey production.

Due to the unbalanced observations, it makes sense that the original correlogram is similar to the non-counterfeit correlogram.

**It's worthwhile also checking across more descriptor labels**

### Mahalanobis:

```{r}
cov(whiskey.num)
```

```{r}
p <- 11
dM <- mahalanobis(whiskey.data[-(1:2)], center=colMeans(whiskey.data[-(1:2)]), cov=cov(whiskey.data[-(1:2)]))

upper.quantiles <- qchisq(c(.9, .95, .99), df=p)
density.at.quantiles <- dchisq(x=upper.quantiles, df=p)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)

ggplot(data.frame(dM), aes(x=dM)) +
	geom_histogram(aes(y=after_stat(density)), bins=8,
				   fill="white", col="black") +
	geom_rug() +
	stat_function(fun=dchisq, args = list(df=p),
				  col="red", size=2, alpha=.7, xlim=c(-5,40)) +
	geom_segment(data=cut.points,
				 aes(x=upper.quantiles, xend=upper.quantiles,
					 y=rep(0,3), yend=density.at.quantiles),
				 col="blue", size=2) +
	xlab("Mahalanobis distances and cut points") +
	ylab("Histogram and density")
```

### PCA

```{r}
whiskey.log <- whiskey.data |> mutate(across(where(is.numeric), log))
whiskey.stand <- whiskey.log |> mutate(across(where(is.numeric), ~(scale(.) %>% as.vector)))

(Sigma <- cov(whiskey.stand[3:13]))
eigen(Sigma)
pca <- as.matrix(whiskey.stand[3:13]) %*% eigen(Sigma)$vectors
```

```{r}
pca <- as.data.frame(pca)
pca$Descriptor <- whiskey.data$Descriptor
summary(pca)

ggplot(pca, aes(x=V1,y=V2, color=factor(Descriptor)))+
  geom_point(size=1)
```

```{r}
pca2 <- prcomp(whiskey.stand[-(1:2)], center=TRUE, scale=TRUE)
summary(pca2)
pca2$rotation
ggbiplot(pca2, obs.scale = 1, var.scale = 1, alpha = .1)
```

```{r}
pca2 <- prcomp(subset(whiskey.stand, Descriptor!="Counterfeit")[,-(1:2)], center=TRUE, scale=TRUE)
plot1 <- ggbiplot(pca2, obs.scale = 1, var.scale = 1, alpha = .1)+
  ylim(-5, 5)+
  ggtitle("Standardized, Log Transformed, Subset of Non-Counterfeit")

pca2 <- prcomp(subset(whiskey.stand, Descriptor=="Counterfeit")[,-(1:2)], center=TRUE, scale=TRUE)
plot2 <- ggbiplot(pca2, obs.scale = 1, var.scale = 1, alpha = .1)+
  ggtitle("Standardized, Log Transformed, Subset of Counterfeit")
#plot2
grid.arrange(plot1, plot2, ncol=2)
```
Once again, similar to the bar plots, immedietly the **Fe** axis jumps out, being negative across PC1.

## Multivariate Normality

```{r}
plot_norm_est <- function(data, parameter, t) {
  ggplot(data) +
    geom_histogram(mapping = aes(x = parameter, y=..density..), fill="steelblue", colour="black", bins = nclass.FD(parameter)) +
    stat_function(fun=dnorm, args= list(mean=0, sd=1), xlim=c(-4, 4)) + 
    ggtitle(t)
}

plot_norm_est(whiskey.stand, whiskey.stand$P, "P")
plot_norm_est(whiskey.stand, whiskey.stand$S, "S")
plot_norm_est(whiskey.stand, whiskey.stand$Cl, "Cl")
plot_norm_est(whiskey.stand, whiskey.stand$K, "K")
plot_norm_est(whiskey.stand, whiskey.stand$Ca, "Ca")
plot_norm_est(whiskey.stand, whiskey.stand$Mn, "Mn")
plot_norm_est(whiskey.stand, whiskey.stand$Fe, "Fe")
plot_norm_est(whiskey.stand, whiskey.stand$Cu, "Cu")
plot_norm_est(whiskey.stand, whiskey.stand$Zn, "Zn")
plot_norm_est(whiskey.stand, whiskey.stand$Br, "Br")
plot_norm_est(whiskey.stand, whiskey.stand$Rb, "Rb")

```

Hardly any of these look normal

```{r}
plot.qq <- function(data, parameter, title) {
  ggplot(data, aes(sample=parameter)) +
    stat_qq(mapping = aes(colour = Descriptor)) + stat_qq_line()+
    ggtitle(paste("Normal QQ Plot for", title, sep=" "))
}
plot.qq(whiskey.stand, whiskey.stand$P, "P")
plot.qq(whiskey.stand, whiskey.stand$S, "S")
plot.qq(whiskey.stand, whiskey.stand$Cl, "Cl")
plot.qq(whiskey.stand, whiskey.stand$K, "K")
plot.qq(whiskey.stand, whiskey.stand$Ca, "Ca")
plot.qq(whiskey.stand, whiskey.stand$Mn, "Mn")
plot.qq(whiskey.stand, whiskey.stand$Fe, "Fe")
plot.qq(whiskey.stand, whiskey.stand$Cu, "Cu")
plot.qq(whiskey.stand, whiskey.stand$Zn, "Zn")
plot.qq(whiskey.stand, whiskey.stand$Br, "Br")
plot.qq(whiskey.stand, whiskey.stand$Rb, "Rb")
```

Again, all of these look skewed to the right. Fe looks the closest

Interestingly you can make out the counterfeits seem to be the strange observations in S.

**TODO: Anderson Darling Test - can't remember the exact name :p**

## Classifiying by distance

```{r}
classify.data <- function(p, drop_i) {
  results <- data.frame(matrix(ncol=4,nrow=nrow(pca)))
  colnames(results) <- c('sample', 'true_class','estimated_class','is_correct')
  for (i in (1:nrow(pca))) {
    if (drop_i) {
      pca.dropped <- pca[-c(i),]
    } else {
      pca.dropped <- pca    
    }
  
    pca.counter <- pca.dropped |> filter(Descriptor == "Counterfeit")
    pca.no.counter <- pca.dropped |> filter(Descriptor != "Counterfeit")
  
    X1 <- pca.counter[(1:p)]
    X2 <- pca.no.counter[(1:p)]
    
    dM.counter <- mahalanobis(pca[i,(1:p)], center=colMeans(X1), cov=cov(X1))
    dM.non.counter <- mahalanobis(pca[i,(1:p)], center=colMeans(X2), cov=cov(X2))
  
    counter <- pchisq(dM.counter, df=p, lower.tail=FALSE)
    non.counter <- pchisq(dM.non.counter, df=p, lower.tail=FALSE)
  
    results[i, ]$sample = i
    if (pca[i,]$Descriptor == as.factor("Counterfeit")) {
      results[i, ]$true_class = "Counterfeit"
      if (counter > non.counter) {
        # correct classifciation
        results[i, ]$estimated_class = "Counterfeit"
        results[i, ]$is_correct = TRUE
      } else {
        # incorrect classification
        results[i, ]$estimated_class = "Non - Counterfeit"
        results[i, ]$is_correct = FALSE
      }
    } else {
      results[i, ]$true_class = "Non - Counterfeit"
      if (non.counter > counter) {
        # correct classification
        results[i, ]$estimated_class = "Non - Counterfeit"
        results[i, ]$is_correct = TRUE
  
      } else {
        # incorrect classification
        results[i, ]$estimated_class = "Counterfeit"
        results[i, ]$is_correct = FALSE
  
      }
    }
  }
  results$true_class <- as.factor(results$true_class)
  results$estimated_class <- as.factor(results$estimated_class)
  results
}
dropped.i <- classify.data(p=2, drop_i=TRUE)
no.dropped.i <- classify.data(p=2, drop_i=FALSE)
```

Building confusion matrix:
```{r}

get.conf.mat <- function(data) {
  conf.mat <- confusionMatrix(data$true_class, data$estimated_class)$table

  ## From https://stackoverflow.com/questions/37897252/plot-confusion-matrix-in-r-using-ggplot
  TClass <- factor(c("counterfeit", "counterfeit", "non-counterfeit", "non-counterfeit"))
  PClass <- factor(c("counterfeit", "non-counterfeit", "counterfeit", "non-counterfeit"))
  Y      <- c(conf.mat[1,1], conf.mat[1,2], 
              conf.mat[2,1], conf.mat[2,2])
  df <- data.frame(TClass, PClass, Y)
  
  ggplot(data =  df, mapping = aes(x = TClass, y = PClass)) +
    geom_tile(aes(fill = Y), colour = "white") +
    geom_text(aes(label = sprintf("%1.0f", Y)), vjust = 1) +
    scale_fill_gradient(low = "blue", high = "red") +
    theme_bw() + theme(legend.position = "none")

}

plot1 <- get.conf.mat(dropped.i) +
  ggtitle("Classifying using Mahalanobis Distance, dropping the ith row")

plot2 <- get.conf.mat(no.dropped.i) +
  ggtitle("Classifying using Mahalanobis Distance, NOT dropping the ith row")

grid.arrange(plot1, plot2, ncol=2)
```
using mahalanobis distance causes a fair bit of error when dropping the ith row for counterfeit observations. Might be because of the poor amount of data?

distance has to be done using PCA, since there would be too many factors compared to the amount of counterfeit observations (p > 5).


